---
title: "easyXpress 24 hours cold trial"
author: "Amanda Peake"
date: "2024-12-17"
output: html_document
editor_options: 
  chunk_output_type: console
---

# R version and packages needed
```{r, include=FALSE}
R.Version()$version.string

#easyXpress
#devtools::install_github("AndersenLab/easyXpress")
library(easyXpress)
library(tidyverse)
library(ggplot2)
library(agricolae)
library(lme4)
library(dplyr)
library(caret)
```


## Load data
```{r, collapse=TRUE}
#Read in data 
cold_24hrs_data <- easyXpress::readXpress(filedir = "C:/Users/amand/OneDrive - University of Toronto/Documents/Academics/PhD/Cold_worms/ImageXpress_assays/24hrscold/24hrscold/Analysis-20241202",
                                          rdafile = "NA_Analysis-20241202.RData",
                                          design = TRUE, 
                                          doseR = FALSE
                                          )


#model selection
ms <- easyXpress::modelSelection(cold_24hrs_data$raw_data)

```


## Flag objects
```{r}
#flag objects that are close to the edge of the well that are difficult to segment properly
ef <- edgeOF(data = ms)

#flag objects that are found within the same primary object which are often debris or improperly segmented worms
cf <- clusterOF(data = ef)


```

Check proportion of flags in each plate
```{r, collapse=TRUE}
#Check flags
c1 <- checkOF(data = cf, strain, notes)
c1$p
```

Check distribution of object length for each plate
```{r, collapse=TRUE}
#visualize size distribution of the ojbect by grouping variables
c2 <- checkObjs(data = cf, OF = "filter", strain, notes)
c2

```

Check if small objects are debris
```{r}
#Add variables that describe the PATH to processed images and well labels
cm <- cf %>%
  dplyr::mutate(i.dir = dplyr::case_when(Metadata_Experiment == "cold24hrs" ~ "C:/Users/amand/OneDrive - University of Toronto/Documents/Academics/PhD/Cold_worms/ImageXpress_assays/24hrscold/24hrscold/Analysis-20241202/processed_images/processed_images/", 
                                         TRUE ~ NA_character_), 
                w.lab = paste(Metadata_Plate, strain, sep = "_"))

#Check models 
cm.out <- checkModels(data = cm, 
                      Metadata_Experiment, Metadata_Plate, 
                      proc.img.dir = "i.dir", 
                     well.label = "w.lab", 
                       out.dir = "C:/Users/amand/OneDrive - University of Toronto/Documents/Academics/PhD/Cold_worms/ImageXpress_assays/24hrscold/24hrscold/Analysis-20241202/checkModels/out/")


```

Remove objects less than 100um and MDHD 
```{r, collapse=TRUE}
#add the user variable that will be converted into a flag
u = cm %>%
  dplyr::mutate(user = dplyr::case_when(worm_length_um < 165 ~ "junk",
                                        model == "MDHD" ~ "junk",
                                        #strain == "QX1211" ~ "junk", #QX1211 wells did not have any worms in them
                                        TRUE ~ NA_character_))

#specify user variable as the flag 
uf <- easyXpress::userOF(data= u, user)

#check data again
checkObjs(data = uf, OF = "filter", strain, notes)

```


Flag objects which are likely debris or improperly segmented worms
```{r}
#cl <- classifierOF(data = uf)
```


Flag objects with extreme worm values
```{r, collapse=TRUE}
#flags objects with extremely large worm_length-um values
#o <- easyXpress::outlierOF(data = cl)
o <- easyXpress::outlierOF(data = uf)

#check objects
checkObjs(data = o, OF = 'filter', strain, notes)

#check propotion of flags
co2 <- checkOF(data = o, strain, notes)
co2
```

View wells to see what data is worth keeping
```{r, collapse=TRUE}
#set seed to select 8 random flagged well with flags
set.seed(99)

#set the flags and filter data
o2 <- easyXpress::setOF(data = o) %>%
  #randomly sample 8 wells
  dplyr::filter(well.id %in% sample(well.id, size = 8))

#make overlay
 vo1 <- easyXpress::viewOverlay(data = o2,
                         proc.img.dir = "i.dir",
                         well.label = "w.lab",
                         obj.label = "model",
                         text.anno = "objectFlag",
                         # save to example dir
                         file = "C:/Users/amand/OneDrive - University of Toronto/Documents/Academics/PhD/Cold_worms/ImageXpress_assays/24hrscold/24hrscold/Analysis-20241202/viewOverlay/overlay.png")
```

remove all flagged object from the data
```{r, collapse=TRUE}
#processed object dataset
proc.objs <- easyXpress::filterOF(o, rmVars = TRUE)
```

## Flag wells

NOTE: did not check for variation between bleaches because all measurements were taken from the same bleach prep. 

Remove flagged objects, summarize data within each well and drop all object related variables from the data to prepare for flagging wells. 
```{r, collapse=TRUE}
raw.wells <- easyXpress::summarizeWells(data = o)
```

Flag wells that have too many or too few objects in them
```{r, collapse=TRUE}
#Make a tf dataframe 
tf <- easyXpress::titerWF(data = raw.wells,
                          Metadata_Experiment, bleach, strain, replicate,
                          doseR = FALSE)

#check the number of objects in each well
n <- easyXpress::nWF(data = tf$d, strain, notes, max = 60, min = 3)
n$p

#flag wells with extreme outlier values
ow <- easyXpress::outlierWF(data = n$d, 
                            Metadata_Experiment, strain, replicate, notes)

#check flagged outlier wells by control vs treatment
cw1 <- easyXpress::checkWF(data = ow, strain, notes)
cw1$p

#check flagged outlier wells be plate
cw2 <- easyXpress::checkWF(data = ow, strain, replicate)
cw2$p
```

Remove all flagged wells from the data
```{r, collapse=TRUE}
#remove wells
fw <- filterWF(data = ow, rmVars = TRUE)

#check balance to see fraction retained after flags are filtered
cb <- checkBalance(data = fw, strain, notes, 
                   design = cold_24hrs_data$design, 
                   x = replicate)
cb$p

#remove QX1211 
drop <- fw %>%
  dplyr::mutate(b.filter =
                  dplyr::case_when(strain == "QX1211" ~  "drop",
                                   TRUE ~ "keep")) %>%
  dplyr::filter(b.filter == "keep") %>%
  dplyr::select(-b.filter)


#check data after removing problematic strains or plates
ce1 <- easyXpress::checkEff(data = drop, strain, 
                            x = notes, 
                            y = median_wormlength_um, 
                            fill = replicate, 
                            scales = "free_x"
                           )
ce1

```


## Finalize results

Account for differences between replicates
```{r, collapse=TRUE}
#regress the effect of independent plates (eg. blocks) for each strain
drop$replicate <- as.character(drop$replicate)

reg <- easyXpress::regEff(data = drop, 
                          d.var = median_wormlength_um, #dependent variable
                          c.var = replicate
                          )

reg$p2
```


Calculate difference between control and treatment. delta() calculates the difference in well summary statistics between the experimental condition and the median control condition within a group. 
```{r, collapse=TRUE}
#Calculate difference
del <- easyXpress::delta(data = reg$d, 
                         replicate, strain, 
                         WF = "filter", 
                        vars = c("median_wormlength_um_reg"),
                         doseR = FALSE)


#write del to csv to be used in manuscript visualisation script
write.csv(del, file = "C:/Users/amand/OneDrive - University of Toronto/Documents/Academics/PhD/Cold_worms/24hours_delta_df.csv", row.names = FALSE)

```


Make a dataframe to compare treatment deltas
```{r}
#Make dataframe
treatment_df <- subset(del, del$notes == "treatment")

##anova
strain_annova <- aov(median_wormlength_um_reg_delta ~ strain + replicate, data = treatment_df)
summary(strain_annova)


#Tukey HSD 
strain_tukey <- HSD.test(strain_annova, trt = "strain")
strain_tukey
```


Calculate broad sense heritability
H^2 = 2751.8/(2751.8+1133.2) = 70.83%
{old value when cl dataframe was not passed to next step as it should have been: H^2 = 4034.1/(2240.5 + 4034.1) = 64.29%}

```{r}
#mixed model 
H_strain_model <- lmer(median_wormlength_um_reg_delta ~ (1|strain), data = treatment_df)
summary(H_strain_model)
vc <- VarCorr(H_strain_model)
print(vc,comp=c("Variance"))


```


Make a textfile to submit to CaeNDR to calculate broad sense heritability
```{r}
#Make a dataframe
H_CaeNDR_df <- data.frame(AssayNumber = treatment_df$replicate, 
                          Strain = treatment_df$strain, 
                          TraitName = treatment_df$Metadata_Experiment, 
                          Replicate = treatment_df$Metadata_Well, 
                          Value = treatment_df$median_wormlength_um_reg_delta)

#Format dataframe to match CaeNDR 
#H_CaeNDR_df$AssayNumber <- as.numeric(sub("p", "", H_CaeNDR_df$AssayNumber))
H_CaeNDR_df$Replicate <- as.numeric(sub("A|B|C|D|E|F|G|H", "", H_CaeNDR_df$Replicate))

#export H_CaeNDR_df to txt file
#write.table(H_CaeNDR_df, file = "C:/Users/amand/OneDrive - University of Toronto/Documents/Academics/PhD/Cold_worms/ImageXpress_assays/24hrscold/24hrscold/heritability_CaeNDR_df.tsv", row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

```



Make a box plot to compare median deltas 

```{r}
#Set order of strains based on sampling latitude 
treatment_df$strain <- factor(treatment_df$strain, levels = c("ECA1286", "DL238", "CB4856", "CX11314", "JU394", "N2", "JU1200"), ordered = TRUE)

#Make dataframe for tukey groupings for the plot 
plot_tukey_groups <- merge(strain_tukey$means, strain_tukey$groups, by = "row.names")


#Box plot comparing median deltas for each strain
treatment_delta_plot <- ggplot(treatment_df, aes(x = strain, y = median_wormlength_um_reg_delta)) + 
  geom_boxplot(outlier.shape = NA) +
  #geom_violin() +
  geom_jitter(shape=16, position=position_jitter(0.2), aes(color = replicate)) + 
  ggtitle("B) 24 hrs at 4°C") + 
  xlab("strain ordered by latitude") + 
  ylab("Δ median length per well") +
  theme_classic() + 
  theme(plot.title = element_text(size=16, face = "bold")) + 
  geom_text(data=plot_tukey_groups, aes(x= Row.names, y = Q75, label = groups), size =5, vjust = -1.5, hjust =-1)
  
treatment_delta_plot

#Box plot not colored by replicates
treatment_delta_plot <- ggplot(treatment_df, aes(x = strain, y = median_wormlength_um_reg_delta)) + 
  geom_boxplot(outlier.shape = NA) +
  #geom_violin() +
  geom_jitter(shape=16, position=position_jitter(0.2)) + 
  ggtitle("B) 24 hrs at 4°C") + 
  xlab("strain ordered by latitude") + 
  ylab("Δ median length per well") +
  theme_classic() + 
  theme(plot.title = element_text(size=16, face = "bold")) + 
  geom_text(data=plot_tukey_groups, aes(x= Row.names, y = Q75, label = groups), size =5, vjust = -2, hjust =-1, color = "blue")
  
treatment_delta_plot



```



## Visualize for comparison for gwas briggsea
```{r}
## Visualize comparison for Briggsae Assay 1
#make control dataframe
control_df <- del %>%
  dplyr::filter(notes == "control")

#make a dataframe of means for all strains in the control treatment
control_means_df <- control_df %>%
  group_by(strain) %>%
  summarise(wormlength_mean_control = mean(median_wormlength_um, na.rm = TRUE))
```



Plot control distribution
```{r}
#make box plot of control plates
control_distribution_plot <- ggplot(control_df, aes(x = reorder(strain, median_wormlength_um, FUN = median) , y = median_wormlength_um)) + 
  geom_boxplot(outlier.shape = NA) +
  #geom_violin() +
  geom_jitter(shape=16, position=position_jitter(0.2), size = 1) + 
  ggtitle("Control Phenotypic distribution elegans 24hrs cold treatment") + 
  xlab("strain") + 
  ylab("median length per well") +
  theme_classic() + 
  theme(plot.title = element_text(size=16, face = "bold")) + 
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 

control_distribution_plot
```


Amanda fooling around testing the delta dataframe
```{r}
#make a test dataframe
test_df <- del

#check if delta is the residual effects subtracting from eachother
test_df$control_reg_minus_well_median_reg <- test_df$median_wormlength_um_reg-test_df$control_median_wormlength_um_reg

#Make a dataframe with the control median lengths for the control
test_df2 <- test_df  %>% 
     dplyr::filter(notes == "control") %>% 
     dplyr::group_by(strain) %>% 
     dplyr::summarise(
     uncorrected_control_mean = median(median_wormlength_um, na.rm = TRUE)) 

#add uncorrected_control_means to test_df
test_df <- dplyr::left_join(test_df, test_df2, by = "strain")

#Calculate delta with residual corrected values
test_df$corrected_control <- test_df$uncorrected_control_mean + test_df$control_median_wormlength_um_reg

test_df$corrected_well_median <- test_df$median_wormlength_um + test_df$median_wormlength_um_reg

test_df$corrected_delta <- test_df$corrected_well_median - test_df$corrected_control


#make treatment dataframe
treatment24hrs_df <- test_df %>%
  dplyr::filter(notes == "treatment")


#run annova
annova_24hrs <- stats::aov(corrected_delta ~ strain, data = treatment24hrs_df)
summary(annova_24hrs)

#run Tukey HSD
tukey_24hrs <- agricolae::HSD.test(annova_24hrs, trt = "strain")
tukey_24hrs

#Make dataframe for tukey groupings for the plot 
plot_tukey_groups24 <- merge(tukey_24hrs$means, tukey_24hrs$groups, by = "row.names")

#Set order of strains based on sampling latitude 
treatment24hrs_df$strain <- factor(treatment24hrs_df$strain)

#set height for tukey labels
plot_tukey_groups24$text_plot_height <- 110

#Box plot
treatment24_delta_plot <- ggplot(treatment24hrs_df, aes(x = strain, y = corrected_delta)) + 
  geom_boxplot(outlier.shape = NA) +
  #geom_violin() +
  geom_jitter(shape=16, position=position_jitter(0.2), size = 1, alpha = 0.5) + 
  xlab("Strain") + 
  ylab("Normalized Animal\nLength (µm)") +
  theme_classic() + 
  geom_text(data=plot_tukey_groups24, 
            aes(x= Row.names, y = text_plot_height, label = groups), 
            size = 3.5, 
            color = "blue") + 
  coord_cartesian(clip = "off") + 
  theme(
    axis.title.y = element_text(size = 10), 
    axis.title.x = element_text(size = 10), 
    axis.text.x = element_text(size = 8)
  )
  
treatment24_delta_plot

```




